Future challenges for Research Software Engineering

### 2025-05-11 future_challenges_for_RSEng.tex
Update future_challenges_for_RSEng.tex
Co-authored-by: Magnus Hagdorn <Magnus.Hagdorn@marsupium.org>
some restructuring

\section{The Situation}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
\begin{story}{Introduction}{intro}
In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
of their future in respective small story boxes.
 \end{story}

\section{Aspects of this future}
It is difficult to structure the different aspects in describing the future. We decided to consider
\begin{itemize}
\item The evolution of the term RSE,
\item How education is reshaped by the digital transformation,
\item What the impact of AI to the work of RSEs will look like,
\item Changes to the Academic Institutions,
\item Ethics and social considerations,
\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s).
\end{itemize}

%\subsection{What do we want to have achieved by 2035}

Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.

With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
% MM-NOTE: if "this is vital" is referring to "software skills" it's not clear -- different paragraphs
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
to research activities in a range of different areas, including the following:
\begin{itemize}
  \item High-quality, robust architecture and design for research software
  \item Seamless integration with research and data infrastructures
  \item Development of software for novel and emerging architectures
  \item Application of specialist numerical and statistical methods within software
  \item Optimisation of codes for use on high-performance computing infrastructure
  \item Green computing - efficient implementations of algorithms/methods and use of hardware
  \item ...
\end{itemize}

@@

\subsection{What is in this future for RSEs?}
FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.

@@

\section{REALLY Long-term(10+ years) aspects}
Technically this would mean taking everything from the previous section for granted and look
at what becomes now possible.

>>>>

\section{Aspects of this future}
%\subsection{What do we want to have achieved by 2035}
Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

So where do RSEs fit into this picture? Even today we expect to see RSEs providing technical skills and input
to research activities in a range of different areas, such as
High-quality, robust architecture and design for research software,
Seamless integration with research and data infrastructures,
Development of software for novel and emerging architectures,
Application of specialist numerical and statistical methods within software,
Optimisation of codes for use on high-performance computing infrastructure,
and Green computing - efficient implementations of algorithms/methods and use of hardware

It is already difficult to structure the spectrum of RSE tasks and responsibilities today (FIXME: Willi Hasselbrings classification paper?)
and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
Nevertheless, we decided to consider
\begin{itemize}
\item The evolution of the term RSE,
\item How education is reshaped by the digital transformation,
\item What the impact of AI to the work of RSEs will look like,
\item Changes to the Academic Institutions,
\item Impact of alternate career paths for RSEs and other RTPs (\emph{Research Technical Professional}s).
\item Ethics and social considerations,
\end{itemize}

\begin{story}{Introduction}{intro}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.\\
In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
of their future in respective small story boxes.
 \end{story}

@@

\section{What is in this future for RSEs and where to go from here?}
FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.

@@
